Widespread nitrous oxide undersaturation in U.S. lakes and reservoirs

### 2023-08-29 manuscript/manuscript_file.Rmd
New figure 6 in manuscript

```{r setup, include=FALSE}
# Specify paths-----
# local path for each user.
localPath <- Sys.getenv("USERPROFILE")
knitr::opts_knit$set(root.dir = "../") # set wd to root of RStudio project

@@

```{r source_scripts, echo=FALSE, warning=FALSE, message=FALSE}
# source scripts------
# some libraries are called within this script.  Not sure if that is necessary
# if libraries are also called above.  Thus far I'm including necessary
# library calls in external script.
# see above for .Rmd working directory
source("manuscript/manuscript_support.R", local = knitr::knit_global())
```

@@

While several measurement campaigns have demonstrated that waterbodies can function as N~2~O sinks[@Soued_etal_2016; @Webb_etal_201; @Kortelainen_etal_2020; @Beaulieu_etal_2015; @Beaulieu_etal_2014; @Musenze_etal_2014], these studies have been conducted at limited spatial scales and encompassed small ranges in potential environmental drivers (e.g. nitrogen loading). Most work on aquatic N~2~O biogeochemistry has focused on streams and rivers, but these systems have been consistently found to be N~2~O sources [@Beaulieu_etal_2008; @Beaulieu_etal_2010; @Quick_etal_2019] and the controls on N~2~O biogeochmistry likely differ between flowing and standing waters. These limitations in data availability have complicated efforts to incorporate N~2~O consumption into freshwater N~2~O budgets.

@@

In this study we present the first national-scale survey of N~2~O concentrations and emissions from CONUS waterbodies. A generalized random-tessellation stratified survey design [@Stevens_Olsen_2004] was used to select a spatially-balanced and representative sample of `r dg %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` **[Roy has 1079. Handpicked and NA sites not used. Update this code.]**sites from CONUS waterbodies that are least 1 hectare in area, at least 1 meter deep, have at least 0.1 hectare of open water, and have a minimum residence time of one week **[(Fig. 1)]**. Sample sites were distributed across all nine major CONUS ecoregions and varied widely in water chemistry, morphometry, watershed land-use, and climate [@epa_data]. `r undersatN2oPercent` percent of the sampled waterbodies were undersaturated in N~2~O and were therefore functioning as N~2~O sinks **[(Fig. 1)]**. We modeled the data using a Bayesian hierarchical approach and predicted the dissolved N~2~O concentration for all 465,897 CONUS waterbodies in the target population. At the population level, `r propN2oSinkMean` percent (95% CI: `r propN2oSinkLower` - `r propN2oSinkUpper`) of CONUS waterbodies were functioning as N~2~O sinks. The Western Mountains ecoregion had the smallest proportion of waterbodies functioning as N~2~O sinks (mean = `r propN2oSinkMeanWM`%, 95% CI: `r propN2oSinkLowerWM` - `r propN2oSinkUpperWM`) and the Northern Plains had the greatest (mean = `r propN2oSinkMeanNPL`%, 95% CI: `r propN2oSinkLowerNPL` - `r propN2oSinkUpperNPL`) **[(Table 1)]**.

@@

## Emission Rates

The emission rates reported here span a broad range (`r min.max$min %>% round(.,2)` - `r min.max$max %>% round(.,2)` mg N~2~O m^-2^ d^-1^), which is expected for this large and diverse collection of waterbodies, but are within the range of values reported in the literature for lentic ecosystems [@Yang_etal_2015; @Kortelainen_etal_2020; @Webb_etal_2019]. Despite widespread N~2~O undersaturation in CONUS waterbodies, the national average N~2~O emission rate was positive (mean = `r national_stats$post_med_en2o`, 95% CI: `r national_stats$LCL_en2o` - `r national_stats$UCL_en2o` mg N~2~O m^-2^ d^-1^) **[(Table 1)]**. This contradiction is due to relatively rare waterbodies with very high emission rates that pull the mean (`r national_stats$post_med_en2o` mg N~2~O m^-2^ d^-1^) to a value that far exceeds the median (`r national_stats$post_median_en2o` mg N~2~O m^-2^ d^-1^) **[(SI Fig. 1)]**. At the ecoregional scale, the Northern Appalachians, Northern Plains, and Upper Midwest ecoregions had mean emission rates less than 0. These ecoregions also had the lowest mean NO~3~ availability of all nine ecoregions **[(Table S1)--is this table updated yet?]**.

@@

>>>>

```{r setup, include=FALSE}
# Specify paths-----
# local path for each user.
localPath <- Sys.getenv("USERPROFILE")
#knitr::opts_knit$set(root.dir = "../") # set wd to root of RStudio project
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file()) # set wd to root of RStudio project

@@

```{r source_scripts, echo=FALSE, warning=FALSE, message=FALSE}
# source scripts------
# some libraries are called within this script.  Not sure if that is necessary
# if libraries are also called above.  Thus far I'm including necessary
# library calls in external script.
# see above for .Rmd working directory
source("manuscript/manuscript_support.R", local = knitr::knit_global())

@@

While several measurement campaigns have demonstrated that waterbodies can function as N~2~O sinks[@Soued_etal_2016; @Webb_etal_2019; @Kortelainen_etal_2020; @Beaulieu_etal_2015; @Beaulieu_etal_2014; @Musenze_etal_2014], these studies have been conducted at limited spatial scales and encompassed small ranges in potential environmental drivers (e.g. nitrogen loading). Most work on aquatic N~2~O biogeochemistry has focused on streams and rivers, but these systems have been consistently found to be N~2~O sources [@Beaulieu_etal_2008; @Beaulieu_etal_2010; @Quick_etal_2019] and the controls on N~2~O biogeochmistry likely differ between flowing and standing waters. These limitations in data availability have complicated efforts to incorporate N~2~O consumption into freshwater N~2~O budgets.

@@

In this study we present the first national-scale survey of N~2~O concentrations and emissions from CONUS waterbodies. A generalized random-tessellation stratified survey design [@Stevens_Olsen_2004] was used to select a spatially-balanced and representative sample of `r dg %>% filter(sitetype == "PROB") %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` sites from CONUS waterbodies that are least 1 hectare in area, at least 1 meter deep, have at least 0.1 hectare of open water, and have a minimum residence time of one week[@EPAdesign] **[(Fig. 1)]**. Sample sites were distributed across all nine major CONUS ecoregions and varied widely in water chemistry, morphometry, watershed land-use, and climate [@epa_data]. `r undersatN2oPercent` percent of the sampled waterbodies were undersaturated in N~2~O and were therefore functioning as N~2~O sinks **[(Fig. 1)]**. We modeled the data using a Bayesian hierarchical approach and predicted the dissolved N~2~O concentration for all 465,897 CONUS waterbodies in the target population. At the population level, `r propN2oSinkMean` percent (95% CI: `r propN2oSinkLower` - `r propN2oSinkUpper`) of CONUS waterbodies were functioning as N~2~O sinks. The Western Mountains ecoregion had the smallest proportion of waterbodies functioning as N~2~O sinks (mean = `r propN2oSinkMeanWM`%, 95% CI: `r propN2oSinkLowerWM` - `r propN2oSinkUpperWM`) and the Northern Plains had the greatest (mean = `r propN2oSinkMeanNPL`%, 95% CI: `r propN2oSinkLowerNPL` - `r propN2oSinkUpperNPL`) **[(Table 1)]**.

@@

## Emission Rates[a]

The emission rates reported here span a broad range (`r min.max$min %>% round(.,2)` - `r min.max$max %>% round(.,2)` mg N~2~O m^-2^ d^-1^), which is expected for this large and diverse collection of waterbodies, but are within the range of values reported in the literature for lentic ecosystems [@Yang_etal_2015; @Kortelainen_etal_2020; @Webb_etal_2019]. Despite widespread N~2~O undersaturation in CONUS waterbodies, the national average N~2~O emission rate was positive (mean = `r national_stats$post_med_en2o`, 95% CI: `r national_stats$LCL_en2o` - `r national_stats$UCL_en2o` mg N~2~O m^-2^ d^-1^) **[(Table 1)]**. This contradiction is due to relatively rare waterbodies with very high emission rates that pull the mean (`r national_stats$post_med_en2o` mg N~2~O m^-2^ d^-1^) to a value that far exceeds the median (`r national_stats$post_median_en2o` mg N~2~O m^-2^ d^-1^) **[(SI Fig. 1)]**.

@@

[a]note overlap between all ecoregion.
all ecoregions overlap 0
Note relationship with lake size.
